The five pillars of computational reproducibility: Bioinformatics and beyond

### 2023-08-11 manuscript/rebuttal.Rmd

---
title: "Response to reviewer comments (The five pillars)"
author: "Mark Ziemann, Pierre Poulain, Anusuiya Bora"
date: "`r Sys.Date()`"
output: word_document
code_folding: hide
fig_width: 7
fig_height: 6
bibliography: 5pillars.bib
csl: plos-computational-biology.csl
---

@@

### Reviewer 1

> In this manuscript, the authors present a framework that computational researchers should follow to maximize the reproducibility of their computational research. The five pillars that they identified are 1) code version control and sharing, 2) compute environment control,  3) literate programming, 4) documentation and 5) FAIR data sharing. They give a fair amount of detail regarding each pillar. This paper contains no original research and is more of an opinion piece. With that in mind, if BiB is interested in this topic, I recommend publishing. The authors make very reasonable suggestions. My only request for edits is to rearrange the section structure of the manuscript to be in line with the numbering of the pillars. Right now it is quite confusing.

Response: We thank the reviewer for their positive feedback.
With regards to the ordering of the five pillars, we have altered Figure 1 so that the pillars now
appear in the same order as they are detailed in sections following.

### Reviewer 2

> Comments to the Author: The article offers a comprehensive perspective on reproducibility in computational research, delivering clear and logically backed recommendations. It is indeed a very interesting topic that needs our attention. Nevertheless, there are areas that could benefit from further development and clarity:

Response: We appreciate reviewer 2's positive feedback and thoughtful suggestions.

> 1. Inclusion of Case Studies or Examples: To strengthen your argument, consider incorporating case studies or practical examples that demonstrate the effectiveness of your recommendations. Highlighting specific instances where these best practices have enhanced reproducibility or solved particular issues could substantiate your points more effectively.

Consider case study of enrichment analysis.

> 2. Expanding on Current Shortcomings: The article mentions that current practices are not meeting the goal of reproducibility. An expanded discussion on the ramifications of these shortcomings on scientific progress could accentuate the urgency and significance of addressing this issue.

After the paragraph ending in "... yet as we see, in practice this is rarely achieved." it would be
good to discuss the ramifications of such shortcomings on scientific progress.
This will highlight the urgency and significance of the proposed measures.

> 3. Focus on Implementation Challenges: While the article acknowledges the importance of certain practices, an increased focus on potential obstacles researchers may face when trying to implement these suggestions, as well as strategies to overcome these challenges, would be insightful.

Just before the paragraph "A lack of technical computer skills in many research groups is another culprit."
We should explicitly discuss implementation challenges.
Part of the preceeding paragraph can also be amended.

> 4. Clear Structure: Consider organizing the article into clearly defined sections such as 'Introduction', 'Background', 'Recommendations', 'Challenges', and 'Conclusion'. This could significantly enhance readability and comprehensibility.

The subheading structure has been amended in line with these recommendations.

* Abstract

* Introduction

* Recommendations

  * End-to-end automated process

  * Literate programming

  * Code version control and persistent sharing

  * Compute environment control

  * FAIR and persistent data sharing

  * Documentation

* Challenges

* Conclusion

* Key points

> 5. Engaging Conclusion: Reworking the conclusion to be more impactful could be beneficial. Instead of merely summarizing the recommendations, it would be useful to indicate the potential trajectory of computational research if these recommendations are adopted, as well as highlighting the potential risks if they aren't.

Response: We have completely rewritten the conclusion.
Instead of reiterating the five pillars, we focus on the motivation for adoption:

>Large scale datasets are becoming commonplace in medicine and life sciences thanks to the commoditisation
>of platform technologies like DNA sequencing, mass spectrometry and advanced imaging systems.
>These big datasets are enabling breakthrough discoveries, but due to their size and complexity, require
>significant expertise to analyze and distill key observations.
>The measures proposed here are designed to enhance transparency and foster the reuse of data and code.
>Such reuse reduces waste and enhances the cost-effectiveness of research.
>The transparency provided by sharing data, code and documentation makes it easier for problems to be
>identified, so that erroneous findings are less likely to pollute the scientific literature.
>Researchers, their institutions, publishers and funders each have a major role to play in fighting the
>reproducibility crisis by encouraging highly reproducible research practices.

### 2023-08-11 manuscript/5pillars.Rmd
working on #26

<strong>Mark Ziemann<sup>1*</sup>, Pierre Poulain<sup>2</sup>, Anusuiya Bora<sup>1</sup></strong>

@@

2. Université Paris Cité, CNRS, Institut Jacques Monod, Paris, France.

@@

Unreliable research is a waste of resources and time and can mislead readers to pursue doomed
research directions.
Much has been written on how irreproducibility has reached crisis levels in psychology and life
sciences [@Baker2015; @Begley2012-di].
Indeed, in bioinformatics, the situation is grim with a 2009 systematic evaluation showing only
2 of 18 articles could be reproduced (11%) @Ioannidis2009-hv, bringing into question the reliability of
those studies.
Workshops conducted by NIH intramural researchers in 2018/2019 sought to reproduce five bioinformatics
studies and could not reproduce *any*, citing missing data, software and documentation
@Zaringhalam2020-ms.
A recent systematic analysis of Jupyter notebooks in biomedical articles had similar observations, with
only 245/4169 notebooks (5.9%) giving similar results compared to the original, with irreproducibility
blamed on missing data, broken dependencies and buggy code @Samuel2022-ha.
A similar survey of R scripts in the Harvard Dataverse repository found slightly better results,
with 26% of scripts completing without errors [@Trisovic2022-dq].
Irreproducible bioinformatics has the potential to place patient safety at risk, as demonstrated by an
earlier case study @Baggerly2010-nr.
Researchers are theoretically able to achieve 100% computational reproducibility, yet as we see,
in practice this is rarely achieved.

@@

## End-to-end automated process

@@

## Literate programming

@@

## Code version control and persistent sharing

@@

## Compute environment control

@@

## FAIR and persistent data sharing

@@

## Documentation

@@

## Limitations and challenges

@@

@@

Reproducibility is an essential feature of reliable computational research yet current practices
are falling short of this goal.
Our recommendations for computational researchers are summarized as:

0. Encapsulate the entire workflow in an end-to-end automated process, executable with a master script.

1. Share the code repository and deposit a snapshot to an archive.

2. Share a controlled environment; options include a Docker or virtual machine image or
configuration for Conda or Guix.

3. Use literate scripting tools like R Markdown or Jupyter notebooks to generate all outputs,
such as charts and tables.

4. Document the workflow sufficiently for reproduction, especially the repository `README`.

5. Share research data and metadata according to FAIR principles.

@@

**Mark Ziemann** is a Senior Lecturer in Biotechnology and Bioinformatics at Deakin University.
His research spans bioinformatics as applied to omics data analysis in biomedicine, as well as aspects
of research rigor.

>>>>

<strong>Mark Ziemann<sup>1,2*</sup>, Pierre Poulain<sup>3</sup>, Anusuiya Bora<sup>1</sup></strong>

@@

2. Burnet Institute, Melbourne, Australia

3. Université Paris Cité, CNRS, Institut Jacques Monod, Paris, France.

@@

Much has been written on how irreproducibility has reached crisis levels in psychology and life
sciences [@Baker2015; @Begley2012-di].
Indeed, in bioinformatics, the situation is grim with a 2009 systematic evaluation showing only
2 of 18 articles could be reproduced (11%) @Ioannidis2009-hv, bringing into question the reliability of
those studies.
Workshops conducted by NIH intramural researchers in 2018/2019 sought to reproduce five bioinformatics
studies and could not reproduce *any*, citing missing data, software and documentation
@Zaringhalam2020-ms.
A recent systematic analysis of Jupyter notebooks in biomedical articles had similar observations, with
only 245/4169 notebooks (5.9%) giving similar results compared to the original, with irreproducibility
blamed on missing data, broken dependencies and buggy code @Samuel2022-ha.
A similar survey of R scripts in the Harvard Dataverse repository found slightly better results,
with 26% of scripts completing without errors [@Trisovic2022-dq].
Irreproducible bioinformatics has the potential to place patient safety at risk, as demonstrated by an
earlier case study @Baggerly2010-nr.
Researchers are theoretically able to achieve 100% computational reproducibility, yet as we see,
in practice this is rarely achieved.

(TODO: discuss the ramifications of such shortcomings on scientific progress.
This will highlight the urgency and significance of the proposed measures.)

(Unreliable research is a waste of resources and time and can mislead readers to pursue doomed
research directions.)

## Recommendations

@@

### End-to-end automated process

@@

### Literate programming

@@

### Code version control and persistent sharing

@@

### Compute environment control

@@

### FAIR and persistent data sharing

@@

### Documentation

@@

## Challenges

@@

(TODO: reviewer 2 point 3: focus on implementation challenges and strategies to overcome.
Expand this section articulating the difficulties with making gold standard reproducibility the norm.
Detail some strategies that can be used.
Case in point: spreadsheet causing gene name errors

1. described in 2004

2. followed by large scale screens in 2016,2022

3. HUGO finally takes action

4. Public awareness

5. Numbers of affected articles diminishes
)

@@

Large scale datasets are becoming commonplace in medicine and life sciences thanks to the commoditisation
of platform technologies like DNA sequencing, mass spectrometry and advanced imaging systems.
These big datasets are enabling breakthrough discoveries, but due to their size and complexity, require
significant expertise to analyze and distill key observations.
The measures proposed here are designed to enhance transparency and foster the reuse of data and code.
Such reuse reduces waste and enhances the cost-effectiveness of research.
The transparency provided by sharing data, code and documentation makes it easier for problems to be
identified, so that erroneous findings are less likely to pollute the scientific literature.
Researchers, their institutions, publishers and funders each have a major role to play in fighting the
reproducibility crisis by encouraging highly reproducible research practices.

@@

**Mark Ziemann** is a Senior Lecturer in Biotechnology and Bioinformatics at Deakin University.
He has recently been recruited to Burnet Institute as Head of Bioinformatics.
His research spans bioinformatics as applied to omics data analysis in biomedicine, as well as aspects
of research rigor.
